Beyond the Numbers: Dissecting New York City's Leading Causes of Death Across Demographics

### 2024-03-17 paper/paper.qmd
Update references in paper

The causes of mortality and determinants of such mortality causes have puzzled researchers for centuries. Despite evidence of the cause of mortality identification inaccuracies {Johns2013}, comprehensive and detailed death certificate records along with more powerful computing and statistical analysis tools have made it much easier to observe actual causes of mortality – at least in the developed world. The determinants of mortality are much harder to pinpoint, with the relative strength of each determinant an even more daunting task.
A popular theory posits that “the wealthier the healthier” – that is, higher incomes directly cause lower mortality, and that higher incomes are the leading cause of lower mortality. Yet this supposed causal relationship can likewise work in reverse in which higher mortality means lower incomes, and studies have shown this explains much of the correlation {Cutler2006}. Surprisingly it has also been observed that recessions seem to improve health outcomes, as people are more likely to do less health-damaging activities (i.e., drinking and smoking) and more health-enhancing activities (i.e., exercise) {Ruhm, 2000}. Another theory suggests that education and ethnicity are the leading causes of lower mortality. For example, according to a study, higher mortality is observed among children of black mothers relative to children of white mothers, regardless of the education level of the mother. However, when comparing only children of white mothers, mothers with higher education levels are observed to show much lower relative mortality for their children {Pamuk1998}. Higher education levels among women are consistently correlated with lower mortality from cardiovascular disease and lung cancer, but also higher mortality from breast cancer {Cutler2006} – demonstrating the importance of comparing mortality by cause. From a long-run perspective, there was hardly any observed difference in life expectancy among England's rich and poor before the Enlightenment period. Yet past this period, there is a widespread general divergence. Since education has always been more accessible for the rich post-Enlightenment, this highlights the strong potential causal link between education (i.e., the development of new ideas of personal and public health such as germ theory of disease, and openness to adjust behaviours or implement policies validating such ideas) and mortality {Cutler2006}. Finally, there is a theory that explains differences in mortality by cause due to differing socio-cultural practices by ethnicity and education levels. For example, lung cancer and cardiovascular diseases should be lower among highly educated people since they are less likely to smoke {Pamuk1998}. While this theory has its merits, the size of its effect is questionable. For example, smoking prevalence is higher among white Americans relative to black Americans, yet this cannot explain black-white differences in American mortality {Cutler2006}.

@@

Lastly, all analysis was conducted using R and various packages {citeR}

@@

This analysis utilizes data from the New York City Department of Health and Mental Hygiene, specifically provided by the Bureau of Vital Statistics {@NYCData}.
The dataset encompasses records of deaths in New York City since 2007, detailing the leading causes of death categorized by sex and ethnicity.
Each entry in this dataset is derived from NYC death certificates which are the official documentation of every death occurring within the city's jurisdiction.

@@

The dataset comprises several key variables, including: - Year: The year of the recorded death.

>>>>

The causes of mortality and determinants of such mortality causes have puzzled researchers for centuries. Despite evidence of the cause of mortality identification inaccuracies [@Johns2013], comprehensive and detailed death certificate records along with more powerful computing and statistical analysis tools have made it much easier to observe actual causes of mortality – at least in the developed world. The determinants of mortality are much harder to pinpoint, with the relative strength of each determinant an even more daunting task.
A popular theory posits that “the wealthier the healthier” – that is, higher incomes directly cause lower mortality, and that higher incomes are the leading cause of lower mortality. Yet this supposed causal relationship can likewise work in reverse in which higher mortality means lower incomes, and studies have shown this explains much of the correlation [@Cutler2006]. Surprisingly it has also been observed that recessions seem to improve health outcomes, as people are more likely to do less health-damaging activities (i.e., drinking and smoking) and more health-enhancing activities (i.e., exercise) [@Ruhm2000]. Another theory suggests that education and ethnicity are the leading causes of lower mortality. For example, according to a study, higher mortality is observed among children of black mothers relative to children of white mothers, regardless of the education level of the mother. However, when comparing only children of white mothers, mothers with higher education levels are observed to show much lower relative mortality for their children [@Pamuk1998]. Higher education levels among women are consistently correlated with lower mortality from cardiovascular disease and lung cancer, but also higher mortality from breast cancer [@Cutler2006] – demonstrating the importance of comparing mortality by cause. From a long-run perspective, there was hardly any observed difference in life expectancy among England's rich and poor before the Enlightenment period. Yet past this period, there is a widespread general divergence. Since education has always been more accessible for the rich post-Enlightenment, this highlights the strong potential causal link between education (i.e., the development of new ideas of personal and public health such as germ theory of disease, and openness to adjust behaviours or implement policies validating such ideas) and mortality [@Cutler2006]. Finally, there is a theory that explains differences in mortality by cause due to differing sociology-cultural practices by ethnicity and education levels. For example, lung cancer and cardiovascular diseases should be lower among highly educated people since they are less likely to smoke [@Pamuk1998]. While this theory has its merits, the size of its effect is questionable. For example, smoking prevalence is higher among white Americans relative to black Americans, yet this cannot explain black-white differences in American mortality [@Cutler2006].

@@

Lastly, all analysis was conducted using R and various packages [@citeR]

@@

This analysis utilizes data from the New York City Department of Health and Mental Hygiene, specifically provided by the Bureau of Vital Statistics [@NYCData].
The data set encompasses records of deaths in New York City since 2007, detailing the leading causes of death categorized by sex and ethnicity.
Each entry in this data set is derived from NYC death certificates which are the official documentation of every death occurring within the city's jurisdiction.

@@

The data set comprises several key variables, including: - Year: The year of the recorded death.
